\center Towards Globally Color-Dual All-Loop Integrands

### 2023-08-17 notes/2loopNLSM.tex
Rearranged subsections in Review slightly

@@

Next we will briefly describe a 2D scalar theory, Zakharov-Mikhailov (ZM) theory, whose equations of motion bear a striking resemblance to the SDYM equatioi of motion given above in \eqn{SDEOM}.
\paragraph{ZM-theory} The Lagrangian for Zakharov-Mikhailov (ZM) theory is a simple $SU(N)$ scalar theory defined in two spacetime dimensions:
\begin{eBox}
\be
\mathcal{L}^{\text{ZM}} = \frac{1}{2}(\partial \varphi)^2 + g f^{abc} \varphi^a (\partial^\mu \varphi^b)( \tilde{\partial}_\mu \varphi^c) =  \text{tr}\left[\frac{1}{2}(\partial \varphi)^2-i g \varphi [\partial_t \varphi, \partial_z \varphi ]\right]
\ee
\end{eBox}
where $\tilde{\partial}_\mu = \epsilon_{\mu\nu}\partial^\nu $ and $\epsilon_{\mu\nu}$ is the two-dimensional Levi-Civita symbol. Expanded out into spacetime components, the momentum space Feynman rule for this theory is simply:
\be
\langle ab\rangle \equiv p_a^0 p_b^1 - p_a^1 p_b^0 = -\langle ba \rangle 
\ee
 From the interaction part of the Lagrangian, we can see that the three-point vertex is (1) functionally symmetric and (2) the kinematics are completely anti-symmetric: 
\be
\cubic{hgrey0}{}{nhpBlue}{}{nhpBlue}{}{nhpBlue} = \langle 12\rangle = \langle 23\rangle = \langle 31\rangle = -\langle 21\rangle
\ee
This important property is due to momentum conservation alone and does not need any additional kinematic constraints to satisfy the duality. In addition to anti-symmetry, the cubic graphs will satisfy jacobi off-shell on any internal edge due to the delta function identity of 2D Levi-Civita symbols:
\be
\langle ab \rangle \langle cd\rangle +\langle ac \rangle \langle db\rangle +\langle ad \rangle \langle bc\rangle = \delta^{[a}_c\delta^{b]}_d + \delta^{[a}_d\delta^{c]}_b+ \delta^{[a}_b\delta^{d]}_c=0
\ee
Theories with this property are the creme de la creme of color-dual gauge theories as they trivially satisfy color-kinematic algebraic relations off-shell. This allows one to trivially promote the duality from tree-level to all-loop order by dressing each cubic vertex with its own kinematic structure constant.

While the color-dual nature of ZM theory is rather elegant, the loop-level construction introduces complications when applying dimensional regularization. Similar to the difficulties of renormalizing chiral fermions \cite{ 'tHooft}, there is ambiguity in promoting the integrands to formal $D$-dimensional expressions. As such, in our two-loop bootstrap of NLSM we will only use the ZM integrands as a mechanism for reducing the residual gauge freedom in our final solution -- rather than a full $D$-dimensional integrand to which we match off-shell in dim reg. 
\paragraph{YZ-Theory} Thus far, the two color-dual Lagrangians we have discussed make explicit reference to the spacetime dimension ($D=4$ for SDYM, and $D=2$ for ZM theory). This makes them poorly suited for constructing dimensionally regulated integrals in loop-level amplitudes. Here we briefly review YZ-theory, which is a honest $D$-dimensional theory first proposed by Cheung and Hsien \cite{Cheung2016prv} for the purpose of describing nonlinear sigma model (NLSM) scattering. One can understand YZ-theory as a particular dimensional reduction of Yang-Mills in $D=2d+1$ dimensions, down to $d$ dimensions.

@@

Both factorization channels are valid descriptions of NLSM amplitudes when plugging in the $Z$ and $\bar{Z}$ on-shell states. As constructed, the $Y$ and $\bar{Z}$ particles may be off-shell, while the $Z$ particles must be placed on shell, $\varepsilon^Z_\mu(k)\rightarrow k_\mu$ in order for these numerators to produce NLSM ordered amplitudes. Given this special property of YZ-theory (namely that the $Z\bar{Z}$ and $YY$ constrctions are equivalent), the $Y$ particle need not be considered when constructing NLSM amplitudes on shell. This is essentially what $J$-theory does, as we will show below.
\paragraph{J-theory} There is an alternative Lagrangian formulation of NLSM constructed by Chueng and one of the authors, that instead starts from the chiral current description of NLSM:

@@

\mathcal{L} = \text{tr}[\partial U \partial U^{-1}] \equiv \text{tr}[J^\mu J_\mu]

@@

where $J_\mu = U\partial_\mu U^{-1}$. To build the J-theory Lagnraigna, we can think of the chiral current $J^\mu$ as a \textit{pure gauge} vector field. Thus, the field strength for $J_\mu$ must vanish, this yielding an equivalent on-shell formulation of the EOM:

@@

\Box J^\mu +g f^{abc} J^\nu J^{\mu\nu} = 0

@@

where we have applied the gauge condition $\partial_\mu J^\mu = 0$. Adding back in a $\bar{J}^\mu$ auxiliary field as a Lagrange multiplier, we obtain the following Lagrangian that is manifestly color-dual:

@@

\mathcal{L}^{\text{J}} = \frac{1}{2}(\partial J)(\partial \bar{J}) +g f^{abc} \bar{J}_{\mu\nu}J^\mu J^\nu

@@

where as above, $\bar{J}_{\mu\nu} = \partial_\mu J_\nu - \partial_\nu J_\mu$. Subject to the Lorenz gauge condition, $\partial_\mu J^\mu =0$. This theory is one-to-one with the observables amplitudes of YZ-theory when only considering $Z-\bar{Z}$ on-shell states. As such, the single vertex in this theory is the same as for YZ theory: 
\cubic{hgrey0}{fermion2}{}{fermion2}{}{fermion2}{} =i(\varepsilon_1 p_{2})(\varepsilon_2\bar{\varepsilon}_3) - i(\varepsilon_2 p_{1})(\varepsilon_1\bar{\varepsilon}_3)

@@

\subsection{Color-dual Bootstrap}
Some background here on how we approach color-dual bootstrap. Building an ansatz from irreducible lorentz products (ISPs). Comments about:
\item loop momenta weight counting 
\item polynomial vector spaces 
\item generalized gauge freedom
\item functional jacobi constraints and cuts
\section{Isolating the cubic sector}
Here we'll describe the why constructing the full Yang-Mills kinematic algebra (outside self-dual sector) is generically hard, and why the same reason makes NLSM beyond one-loop hard as well. It has to do with out the vector state sum mixes with the four-point contact. In terms of our previously defined Feynman rules, the three-point YM vertex takes the form:
\cubic{hgrey0}{gluon}{}{gluon}{}{gluon}{}  =  \cubic{hgrey0}{fermion2}{}{}{nhpRed}{}{nhpRed} + \cubic{hgrey0}{}{nhpRed}{fermion2}{}{}{nhpRed} + \Acubic{hgrey0}{}{nhpRed}{}{nhpRed}{fermion2}{} +\text{h.c.}
The state-sum between two three-point gluon vertex yields the following expansion in terms of YZ-theory vertices:
 \AcubicB{hgrey0}{gluon}{}{gluon}{}{gluon}{}{gluon}{} =  \AcubicB{hgrey0}{}{nhpRed}{}{nhpRed}{fermion2}{}{fermion2}{} + 5 \text{ perms} + \AcubicB{hgrey0}{}{nhpRed}{}{nhpRed}{fermion2}{}{}{nhpRed}
All the $(\epsilon\epsilon)^1$ diagram permutations are isolated to the cubic sector and together manifest color-kinematics duality. However, the exception of the last diagram which introduces a $(\epsilon\epsilon)^2$ factor, must conspire with the Yang-Mills contact to preserve the duality off-shell

@@

@@

\item Write Lagrangian for j-theory and reiterate the argument about why you know it's pions.  Reiterate connection to ZM (dualize j and integrate in phi).  Should probably mention CS somewhere.  Point out that all of the theories that come from have CK are really some version of YM where you (1) impose some condition like F=0 or some other condition on the field strength and (2) you impose Lorenz gauge.

@@

>>>>

\draftnoteJM{I don't think there should be a factor of 1/2 in the kinetic term.}

@@

Next we will briefly describe a 2D scalar theory, Zakharov-Mikhailov (ZM) theory, whose equations of motion bear a striking resemblance to the SDYM equatioi of motion given above in \eqn{SDEOM}.

\paragraph{YZ-Theory} Thus far, the two color-dual Lagrangians we have discussed make explicit reference to the spacetime dimension ($D=4$ for SDYM, and $D=2$ for ZM theory). This makes them poorly suited for constructing dimensionally regulated integrals in loop-level amplitudes. Here we briefly review YZ-theory, which is a honest $D$-dimensional theory first proposed by Cheung and Hsien \cite{Cheung2016prv} for the purpose of describing nonlinear sigma model (NLSM) scattering. One can understand YZ-theory as a particular dimensional reduction of Yang-Mills in $D=2d+1$ dimensions, down to $d$ dimensions.

@@

Both factorization channels are valid descriptions of NLSM amplitudes when plugging in the $Z$ and $\bar{Z}$ on-shell states. As constructed, the $Y$ and $\bar{Z}$ particles may be off-shell, while the $Z$ particles must be placed on shell, $\varepsilon^Z_\mu(k)\rightarrow k_\mu$ in order for these numerators to produce NLSM ordered amplitudes. Given this special property of YZ-theory (namely that the $Z\bar{Z}$ and $YY$ constrctions are equivalent), the $Y$ particle need not be considered when constructing NLSM amplitudes on shell. This is essentially what $J$-thepory does, as we will show below.
\paragraph{J-theory} There is an alternative color-dual formulation of NLSM that the theory manifestly corresponds to NLSM at least at tree level \cite{Cheung:2021zvb}.
For related formulations of NLSM that do not treat color-kinematics duality see \cite{Freedman:1980us, Slavnov:1971mz}.
The construction begins with a first order formulation of NLSM in terms of chiral currents $J^\mu$.
First order formulation of NLSM in terms of chiral current $J^\mu$.
Two equations.
Two constraints, the field strength should vanish

@@

\label{eq:FieldStrengthJ}
F_{\mu\nu}(J) = \partial_\mu J_\nu - \partial_\nu J_\mu + g[J_\mu , J_\nu]=0

@@

which then implies that $J^\mu$ is pure gauge $J_\mu = U \partial_\mu U^{-1}$.
The second condition is that the chiral current should be in Lorenz gauge

@@

\label{eq:LorenzGaugeJ}
\partial_\mu J^\mu=0

@@

which then yields the NLSM equation of motion $\partial_\mu (U \partial^\mu U^{-1})=0$.
Taking a linear combination of \eqref{eq:FieldStrengthJ} and \eqref{eq:LorenzGaugeJ} yields
\Box J^\mu +g f^{abc} J^\nu \partial_\nu J^\mu = 0.
Integrating in an auxiliary field $\bar{J}^\mu$ as a Lagrange multiplier trivially produces the Lagrangian

@@

\mathcal{L}^{\text{J}} =\bar{J}^\mu \Box J_\mu +g f^{abc} \bar{J}^\mu J^\nu \partial_\nu J^\mu .

@@

By calculating an off-shell 4pt correlation function it is possible to check that this Lagrangian automatically satisfies color-kinematics duality.
Kinematic algebra is diff algebra (as usual).
How to extract physical states... (longitudinal J, weird J bar, restores broken Bose).

This theory is one-to-one with the observables amplitudes of YZ-theory when only considering $Z-\bar{Z}$ on-shell states. As such, the single vertex in this theory is the same as for YZ theory: 
\be
\cubic{hgrey0}{fermion2}{}{fermion2}{}{fermion2}{} =i(\varepsilon_1 p_{2})(\varepsilon_2\bar{\varepsilon}_3) - i(\varepsilon_2 p_{1})(\varepsilon_1\bar{\varepsilon}_3)
\ee

\paragraph{ZM-theory}
While J-theory manifests color-kinematics duality, it only does so at tree level because J-theory breaks manifest Bose symmetry.
Manifest color-kinematics duality can be achieved at all loop orders by restricting J-theory to two spacetime dimensions where the chiral current can be dualized to a scalar $J^\mu = \epsilon^{\mu\nu} \partial_\nu \varphi$.
This trivially enforces Lorenz gauge while the vanishing field strength condition...

The Lagrangian for Zakharov-Mikhailov (ZM) theory is a simple $SU(N)$ scalar theory defined in two spacetime dimensions:
\begin{eBox}
\be
\mathcal{L}^{\text{ZM}} = \frac{1}{2}(\partial \varphi)^2 + g f^{abc} \varphi^a (\partial^\mu \varphi^b)( \tilde{\partial}_\mu \varphi^c) =  \text{tr}\left[\frac{1}{2}(\partial \varphi)^2-i g \varphi [\partial_t \varphi, \partial_z \varphi ]\right]
\ee
\end{eBox}
where $\tilde{\partial}_\mu = \epsilon_{\mu\nu}\partial^\nu $ and $\epsilon_{\mu\nu}$ is the two-dimensional Levi-Civita symbol. Expanded out into spacetime components, the momentum space Feynman rule for this theory is simply:
\be
\langle ab\rangle \equiv p_a^0 p_b^1 - p_a^1 p_b^0 = -\langle ba \rangle 
\ee
 From the interaction part of the Lagrangian, we can see that the three-point vertex is (1) functionally symmetric and (2) the kinematics are completely anti-symmetric: 
\be
\cubic{hgrey0}{}{nhpBlue}{}{nhpBlue}{}{nhpBlue} = \langle 12\rangle = \langle 23\rangle = \langle 31\rangle = -\langle 21\rangle
\ee
This important property is due to momentum conservation alone and does not need any additional kinematic constraints to satisfy the duality. In addition to anti-symmetry, the cubic graphs will satisfy jacobi off-shell on any internal edge due to the delta function identity of 2D Levi-Civita symbols (TODO:  show the identity, say that the following is Schouten, cite Schouten?, comment on addition of masses):
\be
\langle ab \rangle \langle cd\rangle +\langle ac \rangle \langle db\rangle +\langle ad \rangle \langle bc\rangle = \delta^{[a}_c\delta^{b]}_d + \delta^{[a}_d\delta^{c]}_b+ \delta^{[a}_b\delta^{d]}_c=0
\ee
Theories with this property are the creme de la creme of color-dual gauge theories as they trivially satisfy color-kinematic algebraic relations off-shell. This allows one to trivially promote the duality from tree-level to all-loop order by dressing each cubic vertex with its own kinematic structure constant.

While the color-dual nature of ZM theory is rather elegant, the loop-level construction introduces complications when applying dimensional regularization. Similar to the difficulties of renormalizing chiral fermions \cite{ 'tHooft}, there is ambiguity in promoting the integrands to formal $D$-dimensional expressions. As such, in our two-loop bootstrap of NLSM we will only use the ZM integrands as a mechanism for reducing the residual gauge freedom in our final solution -- rather than a full $D$-dimensional integrand to which we match off-shell in dim reg.

@@

\subsection{Color-dual Bootstrap}
Some background here on how we approach color-dual bootstrap. Building an ansatz from irreducible lorentz products (ISPs). Comments about:
\begin{itemize}
\item loop momenta weight counting 
\item polynomial vector spaces 
\item generalized gauge freedom
\item functional jacobi constraints and cuts
\end{itemize} 
\section{Isolating the cubic sector}
Here we'll describe the why constructing the full Yang-Mills kinematic algebra (outside self-dual sector) is generically hard, and why the same reason makes NLSM beyond one-loop hard as well. It has to do with out the vector state sum mixes with the four-point contact. In terms of our previously defined Feynman rules, the three-point YM vertex takes the form:
\be
\cubic{hgrey0}{gluon}{}{gluon}{}{gluon}{}  =  \cubic{hgrey0}{fermion2}{}{}{nhpRed}{}{nhpRed} + \cubic{hgrey0}{}{nhpRed}{fermion2}{}{}{nhpRed} + \Acubic{hgrey0}{}{nhpRed}{}{nhpRed}{fermion2}{} +\text{h.c.}
\ee
The state-sum between two three-point gluon vertex yields the following expansion in terms of YZ-theory vertices:
\be
 \AcubicB{hgrey0}{gluon}{}{gluon}{}{gluon}{}{gluon}{} =  \AcubicB{hgrey0}{}{nhpRed}{}{nhpRed}{fermion2}{}{fermion2}{} + 5 \text{ perms} + \AcubicB{hgrey0}{}{nhpRed}{}{nhpRed}{fermion2}{}{}{nhpRed}
\ee
All the $(\epsilon\epsilon)^1$ diagram permutations are isolated to the cubic sector and together manifest color-kinematics duality. However, the exception of the last diagram which introduces a $(\epsilon\epsilon)^2$ factor, must conspire with the Yang-Mills contact to preserve the duality off-shell

@@

\item Mention NANS after j theory.  Color kinematics duality doesn't care about the propagator but using a funny one might mess up locality in the double copied theory.

@@

\item Write Lagrangian for j-theory and reiterate the argument about why you know it's pions.  Reiterate connection to ZM (dualize j and integrate in phi).  Should probably mention CS somewhere.  Point out that all of the theories that come from have CK are really some version of YM where you (1) impose some condition like F=0 or some other condition on the field strength and (2) you impose Lorenz gauge.  Point out that the kinematic algebra is always diffs.
\item Cite papers arguing about whether NLSM in 2D is non-trivial or not.

@@

For 4pt on-shell ZM, the LRLR configuration is zero and for every other configuration of external states the on-shell amplitude is subtle because of IR poles.  Note that the (on-shell) kinematic numerators are always unambiguous and never suffer from any of these subtleties in any kinematic configuration.  So when you compare against ZM, you either comparing against zero or something subtle.  This is why you can't really take cuts of the NLSM and ZM integrands and meaningfully compare them even though you can compare their numerators.
